Efficient pedigree recording for fast population genetics simulation

### 2017-11-18 forwards_paper.tex
edited edits

However, it is thought that most genetic variation is selectively neutral (or nearly so).
By definition, the alleles carried by individuals in a population at neutral sites
do not affect the population process.
For this reason, if one records the entire genealogical history of a population over the course of a simulation,
one can lay down neutral mutations on top of that history afterwards.
\jda{WLOG seems unclear/ not needed}
Precisely, we would need to know the genealogical tree relating all sampled individuals
at each position along the genome.
In this paper, describe algorithmic tools and data structures developed for the 
coalescent simulator \msprime{}
to efficiently record, and later process, this history.

@@

\jda{Is this the right place for this paragraph? should it maybe be integrated
	with graf beginning 'Modern computing' above?}
\plr{this paragraph is supposed to cite the fwds time simulation methods we need to - @molpopgen fix this up?}
Forwards-time population genetics simulations are growing in importance,
as the genome-wide impacts of hitchhiking and background selection are more widely appreciated.
For instance, \citet{harris2016genetic} used SLiM \citep{haller2017flexible} to simulate tens of thousands of human exomes
to assess the impact of genetic load and Neanderthal introgression on human genetic diversity.
This is roughly state-of-the-art 
-- the fastest simulation frameworks such as SLiM and fwdpp \citep{fwdpp}
can ``only'' simulate tens of megabases of sequence in tens of thousands of individuals
for tens of thousands of generations --
but still substantially short of whole genomes for most biological population sizes.
\jda{I'd find this more compelling with something added like: "e.g., Xs of
	gigabases for Ys of individuals, implying 10Ys of generations needed to
	equilibrate}
There has been some progress at parallelization \citep{lawrie2017accelerating},
which has substantial promise.
\jda{this is vague, promise for what?}
The general idea of storing genealogical information to speed up simulations is not new --
it was entirely implemented in AnA-FiTS \citep{aberer2013rapid}, for instance --
but this program did not implement the critical step of discarding irrelevant genealogical information.
A similar but more limited method for discarding this information does appear in \citet{padhukasahasram2008exploring}.

@@

Below, we first describe the conceptual and algorithmic foundations for the method:
(a) a format, implemented in the \msprime{} Python API,
for storing tree sequences in several \emph{tables};
(b) an algorithm, \algref{W}, to record tree sequence information into these tables on the fly
    from a forwards-time simulation;
and (c) a method to \emph{simplify} a tree sequence, i.e., remove redundant information.
We then describe a method that combines (b) and (d) to allows efficiently record simplified a tree sequence,
and present analyses of its run time complexity.
This is followed by benchmarking comparisons between two implementations of our method,
each using previously published forwards-time simulation libraries.

%%%%%%
\subsection*{Storing a tree sequence}

@@

To record the genealogical history of a forwards in time simulation
we need to record two things for each new chromosome:
the birth time, and the endpoints and parental IDs of each distinctly inherited segment,
which are naturally stored as the \emph{nodes} and \emph{edges} of a tree sequence.
For concreteness, here we write out in pseudocode how to run a neutral Wright--Fisher simulation
(but with overlapping generations) that records genealogical history in this way.
The simulation will run for $T$ generations,
and has $N$ haploid individuals, each carrying a single chromosome of length L.
For simplicity we sample exactly one crossover per generation.
The probability of death per individual each generation is $\delta$.
\jda{In the below we are using the integer vector encoding of a tree as in
	\msprime{} should say so and reference it as the role of node ids is
	otherwise more confusing than it need be}
We will use
$\randomuniform(A)$ to denote an element of the set $A$ chosen uniformly at random 
(and each such instance represents and independent draw).

@@

\algstep{W4.}{Mortality.}{ If $\randomuniform([0, 1)) \geq \delta$ go to \algref{W8}.
}

@@

\algstep{W6.}{Choose parents.}{Set $a \leftarrow \randomuniform(\{0, \dots, n - 1\})$,
    $b \leftarrow \randomuniform(\{0, \dots, n - 1\})$ and $x \leftarrow \randomuniform([0, L))$.
}

@@

\algstep{W8.}{Loop tail}{Set $j \leftarrow j + 1$ and go to \algref{W3}.
}

@@

Algorithm~\algref{W} is a very simple implementation of a Wright--Fisher model
intended to illustrate how we can record the nodes and edges of a tree sequence
forwards in time. We begin in~\algref{W1} by allocating our initial population $P$
and creating $N$ nodes with birth time $T$ generations ago, recorded in the
vector $\tau$. The set $E$ is used to store the edges that we output during the
simulation, and $n$ is the number of nodes created so far (and so, the ID of
the next node we create). Steps~\algref{W2} and~\algref{W3} simply loop over
the generation clock $t$ and individual index $j$. In~\algref{W4} we check if
an individual $P_j$ has died in this generation. If it has, we replace it in
steps \algref{W5}--\algref{W7}; if not, we proceed immediately to the next
individual. When an individual in the population dies, we first allocate
a new node with ID $n$ in~\algref{W5} and record its birth time. Then,
in step~\algref{W6} we choose two indexes $a$ and $b$ uniformly (giving us
parents $P_a$ and $P_b$) and choose a breakpoint $x$. We record the effects of this
event by storing two new edges: one recording that the parent of node $n$
from $0$ to $x$ is $P_a$, and another recording that the parent of $n$
from $x$ to $L$ is $P_b$. We then complete the replacement event by incrementing
$n$, ready to represent the next new node.

@@

The approach that we take is based on Hudson's algorithm for simulating
the coalescent with recombination~\citep{hudson1983properties,kelleher2016efficient};
the implementaiton of simplify parallels the implementation of Hudson's algorithm in \msprime.
Conceptually, this works by 
(a) beginning by painting the chromosome each sample a distinct color;
(b) moving back through history, 
copying the colors of each chromosome to the portions of its' parental chromosomes
from which it was inherited;
(c) each time we would paint two colors in the same spot (a coalescence),
record that information as an edge and instead paint a brand-new color; 
and
(d) once all colors have coalesced on a given segment,
stop propagating it.
This ``paint pot'' description misses some details --
for instance, we must ensure that all coalescing segments in a given individual
are assigned the \emph{same} new color --
but is reasonably close.
The process is depicted in Figure~\ref{fig:simplify_state}.
\jda{Need refs to Figure~\ref{fig:method_diagram} in the text}

@@

\subsection*{Forward simulation and sequential simplification to record simplifed tree sequences}

At any point in Algorithm~\algref{W}, genealogical history is recorded
in an unsimplified tree sequence. This has two additional advantages. First, simplification
can be run periodically through the simulation, taking the set of samples to be
the entire currently alive population. This is important as it keeps memory
usage from growing linearly (and quickly) with time. Second, the simulation can
be \emph{begun} with a tree sequence produced by some other method -- for
instance, by a coalescent simulation with \msprime. 
This process of forward simulation coupled with sequential simplification is our
method

@@

For efficiency, \msprime{} makes several other sortedness requirements on the tables,
that are not necessarily satisfied by tables emitted by a forwards-time simulation.
Fortunately, \msprime{}'s API includes tools to rectify this by first sorting (using \texttt{sort\_tables})
and then using \texttt{simplify}, which works on sorted tables
and is guarenteed to produce a valid, \msprime{}-ready tree sequence.

>>>>

However, it is thought that most genetic variation is selectively neutral (or nearly so).
By definition, the alleles carried by individuals in a population at neutral sites
do not affect the population process.
For this reason, if one records the entire genealogical history of a population over the course of a simulation,
simply laying down neutral mutations on top of that history afterwards
is equivalent to having generated them during the simulation.
Precisely, we would need to know the genealogical tree relating all sampled individuals
at each position along the genome.
In this paper, we describe how algorithmic tools and data structures developed for the 
coalescent simulator \msprime{}
can be used to efficiently record, and later process, this history.

@@

\jda{Is this the right place for this paragraph? should it maybe be integrated
	with graf beginning 'Modern computing' above?}
\plr{this paragraph is supposed to cite the fwds time simulation methods we need to - @molpopgen fix this up?}
Forwards-time population genetics simulations are growing in importance,
as the genome-wide impacts of hitchhiking and background selection are more widely appreciated.
For instance, \citet{harris2016genetic} used SLiM \citep{haller2017flexible} to simulate tens of thousands of human exomes
to assess the impact of genetic load and Neanderthal introgression on human genetic diversity.
This is roughly state-of-the-art 
-- the fastest simulation frameworks such as SLiM and fwdpp \citep{fwdpp}
can ``only'' simulate tens of megabases of sequence in tens of thousands of individuals
for tens of thousands of generations.
This is still substantially short of whole genomes (hundreds to thousands of megabases)
for many biological population sizes of interest 
(e.g., millions for a restricted population of mosquitos).
Achieving these sizes will likely require parallelization \citep{lawrie2017accelerating}.
The general idea of storing genealogical information to speed up simulations is not new --
it was entirely implemented in AnA-FiTS \citep{aberer2013rapid}, for instance --
but this program did not implement the critical step of discarding irrelevant genealogical information.
A similar but more limited method for discarding this information does appear in \citet{padhukasahasram2008exploring}.

@@

Below, we first describe the conceptual and algorithmic foundations for the method:
(a) a format, implemented in the \msprime{} Python API,
for recording tree sequences in several \emph{tables};
(b) an algorithm, \algref{W}, to record tree sequence information into these tables on the fly
    from a forwards-time simulation;
and (c) an algorithm to \emph{simplify} a tree sequence, i.e., remove redundant information.
We then describe a general-purpose method that uses these tools
(as implemented in \msprime{}) 
to efficiently record simplified a tree sequence,
and analyze of its run time complexity.
This is followed by benchmarking the performance improvement achieved by connecting
two previously published forwards-time simulation libraries,
\simupop{} \citep{simupop} and \fwdpp{} \citep{fwdpp}
to the tools in \msprime{}.

%%%%%%
\subsection*{Recording a tree sequence}

@@

To record the genealogical history of a forwards in time simulation
we need to record two things for each new chromosome:
the birth time, and the endpoints and parental IDs of each distinctly inherited segment,
which are naturally stored as the \emph{nodes} and \emph{edges} of a tree sequence.
For concreteness, here we write out in pseudocode how to run a neutral Wright--Fisher simulation
(but with overlapping generations) that records genealogical history in this way.
The simulation will run for $T$ generations,
and has $N$ haploid individuals, each carrying a single chromosome of length L.
For simplicity we sample exactly one crossover per generation.
The probability of death per individual each generation is $\delta$.
We will use
$\randomuniform(A)$ to denote an element of the set $A$ chosen uniformly at random 
(and each such instance represents and independent draw).

We begin in~\algref{W1} by allocating our initial population $P$
(a vector of $N$ node IDs),
and creating $N$ nodes with birth time $T$ generations ago, recorded in the
vector $\tau$. The set $E$ is used to store the edges that we output during the
simulation, and $n$ is the number of nodes created so far (and so, the ID of
the next node we create). Steps~\algref{W2} and~\algref{W3} simply loop over
the generation clock $t$ and individual index $j$. In~\algref{W4} we check if
an individual $P_j$ has died in this generation. If it has, we replace it in
steps \algref{W5}--\algref{W7}; if not, we proceed immediately to the next
individual. When an individual in the population dies, we first allocate
a new node with ID $n$ in~\algref{W5} and record its birth time. Then,
in step~\algref{W6} we choose two indexes $a$ and $b$ uniformly (giving us
parental IDs $P_a$ and $P_b$) and choose a breakpoint $x$. We record the effects of this
event by storing two new edges: one recording that the parent of node $n$
from $0$ to $x$ is $P_a$, and another recording that the parent of $n$
from $x$ to $L$ is $P_b$. We then complete the replacement event by incrementing
$n$, ready to represent the next new node.

@@

\algstep{W4.}{Mortality.}{ If $\randomuniform([0, 1)) \geq \delta$ go to \algref{W9}.
}

@@

\algstep{W6.}{Choose parents.}{Set $a \leftarrow \randomuniform(\{0, \dots, N - 1\})$,
    $b \leftarrow \randomuniform(\{0, \dots, N - 1\})$ and $x \leftarrow \randomuniform([0, L))$.
}

@@

\algstep{W9.}{Loop tail}{Set $j \leftarrow j + 1$ and go to \algref{W3}.
}

@@

@@

The approach that we take is based on Hudson's algorithm for simulating
the coalescent with recombination~\citep{hudson1983properties,kelleher2016efficient};
the implementaiton of simplify parallels the implementation of Hudson's algorithm in \msprime.
Conceptually, this works by 
(a) beginning by painting the chromosome each sample a distinct color;
(b) moving back through history, 
copying the colors of each chromosome to the portions of its' parental chromosomes
from which it was inherited;
(c) each time we would paint two colors in the same spot (a coalescence),
record that information as an edge and instead paint a brand-new color; 
and
(d) once all colors have coalesced on a given segment,
stop propagating it.
This ``paint pot'' description misses some details --
for instance, we must ensure that all coalescing segments in a given individual
are assigned the \emph{same} new color --
but is reasonably close.
The process is depicted in Figures~\ref{fig:method_diagram} and \ref{fig:simplify_state}.

@@

\subsection*{Forward simulation and sequential simplification}

At any point in a simulation scheme like Algorithm~\algref{W}, genealogical history is recorded
in a tree sequence. This has two additional advantages. First, simplification
can be run periodically through the simulation, taking the set of samples to be
the entire currently alive population. This is important as it keeps memory
usage from growing linearly (and quickly) with time. Second, the simulation can
be \emph{begun} with a tree sequence produced by some other method -- for
instance, by a coalescent simulation with \msprime. 
This process of forward simulation coupled with sequential simplification is our
method

@@

For efficiency, \msprime{} makes several other sortedness requirements on the tables,
that are not necessarily satisfied by tables emitted by a forwards-time simulation.
\msprime{}'s API includes tools to rectify this by first sorting (using \texttt{sort\_tables})
and then using \texttt{simplify}, which works on sorted tables
and is guarenteed to produce a valid, \msprime{}-ready tree sequence.
